---
title: "openstatsguide"
subtitle: "Minimum Viable Good Practices for High Quality Statistical Software Packages"
author: "0.0-2"
author-title: "Version"
date: "31 May 2024"
bibliography: references.bib
nocite: |
  @*
csl: computational-statistics.csl
---

*Note: This is a beta release of "openstatsguide" to accompany the useR 2024 abstract submission. It is not yet available interactively via the openstatsware.org website and will undergo further expert input before publication of the first release version 0.1.*

# Scope

We encourage developers of statistical software packages to follow this minimum set of good practices around:

> "**D**ocumentation, **V**ignettes, **T**ests, **F**unctions, **S**tyle, **L**ife cycle"

These keywords can be easily remembered with the mnemonic bridge sentence:

> "**D**evelopers **V**alue **T**ests **F**or **S**oftware **L**ongevity"

While the recommendations are rather generic, we focus on functional programming languages and give links to implementations in R, Python and Julia.

This guide primarily addresses developers of statistical packages. Users interested in assessing the quality of existing statistical packages will find complementary "validation" focused resources valuable, as listed in [References](#references).

# Recommendations

```{r echo=FALSE}
glue_or_drop <- function(doc, img) {
    if (identical(doc, "")) {
        NULL
    } else {
        htmltools::tags$td(
            htmltools::tags$details(
                htmltools::tags$summary(
                    htmltools::tags$img(src = img, height = 20)
                ),
                doc
            )
        )
    }
}
guide_table <- function(r = "", python = "", julia = "") {
    contents <- htmltools::tags$html(
        glue_or_drop(r, "https://www.r-project.org/logo/Rlogo.svg"),
        glue_or_drop(python, "https://s3.dualstack.us-east-2.amazonaws.com/pythondotorg-assets/media/files/python-logo-only.svg"),
        glue_or_drop(julia, "https://raw.githubusercontent.com/JuliaLang/julia-logo-graphics/master/images/julia-dots.svg"),
        sep = "\n"
    )
    htmltools::tags$table(
        style = htmltools::css(
            width = "100%"
        ),
        htmltools::tags$tbody(contents)
    )
}
```

## Design 

Before you begin coding it is often valuable to spend time designing and planning your package in collaboration with the target user. This can also help you define the scope of your first release ensuring that you are working on high-value features and enabling developers to focus. 

::: {.panel-tabset}

## Wireframing

Wireframing involves creating a rough drawing of the user-interface of a webpage or application. Popular tools include:  
[figma](https://www.figma.com/design/)


## Diagrams
Unified Modeling Language (UML)is a way to visually represent the architecture, structure and behaviour or your package. UML diagrams can be drawn programmically in tools such as:   
[mermaid](https://mermaid.js.org/syntax/classDiagram.html)


## Project Management
Defining your minimum viable product, i.e. the scope of your first release will help developers focus and also give an opportunity to validate your understanding of the problem with your target user group.   
[github](https://github.com/features/issues)
[JIRA](https://www.atlassian.com/software/jira)

:::

## Documentation

Documentation is important for both users and developers to understand all objects in your package, without reading and interpreting the underlying source code.

1.  Use **in-line comments** next to functions, classes and other objects to generate their corresponding documentation.

    ```{r results="asis", echo=FALSE}
guide_table(
    r = htmltools::a("roxygen2", href = "https://roxygen2.r-lib.org/"),
    python = htmltools::a("docstrings", href = "https://peps.python.org/pep-0257/"),
    julia = htmltools::a("docstrings", href = "https://docs.julialang.org/en/v1/manual/documentation/")
)
    ```

::: {.panel-tabset}

## <img src="https://www.r-project.org/logo/Rlogo.svg" alt="alt text" width="25">

[roxygen2](https://roxygen2.r-lib.org/)

## <img src="https://s3.dualstack.us-east-2.amazonaws.com/pythondotorg-assets/media/files/python-logo-only.svg" alt="alt text" width="25">

[docstrings](https://peps.python.org/pep-0257/)

## <img src="https://raw.githubusercontent.com/JuliaLang/julia-logo-graphics/master/images/julia-dots.svg" width="25">

[docstrings](https://docs.julialang.org/en/v1/manual/documentation/)

:::

2.  Do also **document internal functions** and classes for maintenance by future developers.

3.  Add **code comments** for ambiguous or complex pieces of internal code, but only after optimizing the code design as much as possible.

## Vignettes

Vignettes are documents that complement the object documentation by providing a comprehensive and long-form overview of your package's functionality from a user point of view, with particular emphasis on the connection to the statistical approaches.

1.  Provide an **introduction vignette** that introduces the package to new users, such that they have an easy entry point for getting started. Make sure to include code examples and automatically compile the vignette to ensure reproducibility.

2.  Include **deep dive vignettes** that go into depth on specific use cases, functionalities or underlying theory, in particular describing the underlying statistical methodology and how it is implemented in the package.

3.  Host your vignettes on a **dedicated website**, which allows users to read the vignettes without installing the package, and simplifies citing the vignettes in other publications.

    ```{r results="asis", echo=FALSE}
guide_table(
    r = htmltools::a("pkgdown", href = "https://pkgdown.r-lib.org/"),
    python = htmltools::a("Sphinx", href = "https://www.sphinx-doc.org/en/master/"),
    julia = htmltools::a("Documenter", href = "https://documenter.juliadocs.org/")
)
    ```

::: {.panel-tabset}

## <img src="https://www.r-project.org/logo/Rlogo.svg" alt="alt text" width="25">

[pkgdown](https://pkgdown.r-lib.org/)

## <img src="https://s3.dualstack.us-east-2.amazonaws.com/pythondotorg-assets/media/files/python-logo-only.svg" alt="alt text" width="25">

[Sphinx](https://www.sphinx-doc.org/en/master/)

## <img src="https://raw.githubusercontent.com/JuliaLang/julia-logo-graphics/master/images/julia-dots.svg" width="25">

[Documenter](https://documenter.juliadocs.org/)

:::

## Tests

Tests are a fundamental safety net and development tool to ensure that your package works as expected, both during development as well as on user systems.

1.  Write **unit tests** for all functions and classes in your package, to ensure that all building blocks work correctly on their own ("white box" testing). Expect to rewrite tests for internal code when you refactor it.

2.  Write **functional tests** for all user facing functionality ("black box" testing). These tests ensure that the user API is stable when refactoring internal code.

    ```{r results="asis", echo=FALSE}
guide_table(
    r = htmltools::a("testthat", href = "https://testthat.r-lib.org/"),
    python = htmltools::a("pytest", href = "https://pytest.org/"),
    julia = htmltools::a("Test", href = "https://docs.julialang.org/en/v1/stdlib/Test/")
)
    ```

::: {.panel-tabset}

## <img src="https://www.r-project.org/logo/Rlogo.svg" alt="alt text" width="25">

[testthat](https://testthat.r-lib.org/)

## <img src="https://s3.dualstack.us-east-2.amazonaws.com/pythondotorg-assets/media/files/python-logo-only.svg" alt="alt text" width="25">

[pytest](https://pytest.org/)

## <img src="https://raw.githubusercontent.com/JuliaLang/julia-logo-graphics/master/images/julia-dots.svg" width="25">

[Test](https://docs.julialang.org/en/v1/stdlib/Test/)

:::   

3.  In addition, ensure a **good coverage** of your code with your tests as a final check, but only after having unit and functional tests on all levels of the code.

    ```{r results="asis", echo=FALSE}
guide_table(
    r = htmltools::a("covr", href = "https://covr.r-lib.org/"),
    python = htmltools::a("Coverage.py", href = "https://coverage.readthedocs.io/"),
    julia = htmltools::a("Coverage.jl", href = "https://github.com/JuliaCI/Coverage.jl")
)
    ```

## Functions

Function definitions should be short, simple and enforce argument types with assertions.

1.  Write **short functions** with less than 50 lines of code for a single and well-defined purpose, with **few arguments**, and low cyclomatic complexity, in order to reduce the risk of bugs, simplify writing tests and ensure that you can maintain them.

2.  Use **type hints** or types to explain to the user which argument of the function expects which type of input.

    ```{r results="asis", echo=FALSE}
guide_table(
    r = htmltools::a("roxytypes", href = "https://openpharma.github.io/roxytypes/"),
    python = htmltools::a("typing", href = "https://docs.python.org/3/library/typing.html"),
    julia = htmltools::a("types", href = "https://docs.julialang.org/en/v1/manual/types/")
)
    ```

3.  Enforce types and other expected properties of function arguments with **assertions**, which give an early and readable error message to the user instead of failing function code downstream in a less explicable way.

    ```{r results="asis", echo=FALSE}
guide_table(
    r = htmltools::a("checkmate", href = "https://mllg.github.io/checkmate/"),
    python = htmltools::a("assertpy", href = "https://pypi.org/project/assertpy/"),
    julia = htmltools::a("ArgCheck.jl", href = "https://github.com/jw3126/ArgCheck.jl")
)
    ```

## Dependency Management
Packages will typically reference other packages to import pre-existing functionality. These dependencies are moving targets, i.e. they are periodically updated. Therefore package dependencies should be proactively managed. 

::: {.panel-tabset}

## <img src="https://www.r-project.org/logo/Rlogo.svg" alt="alt text" width="25">

[renv](https://cran.r-project.org/web/packages/renv/index.html)

## <img src="https://s3.dualstack.us-east-2.amazonaws.com/pythondotorg-assets/media/files/python-logo-only.svg" alt="alt text" width="25">

[conda](https://conda.io/projects/conda/en/latest/index.html)

## <img src="https://raw.githubusercontent.com/JuliaLang/julia-logo-graphics/master/images/julia-dots.svg" width="25">

[Pkg.jl](https://pkgdocs.julialang.org/v1/)

:::

## Style

A consistent and readable code style that is language idiomatic should be used and enforced by style checks.

1.  Use **language idiomatic** code and follow the **"clean code" rules** (use descriptive and meaningful names, prefer simpler over more complex code, avoid duplication of code, regularly refactor code), while allowing for exceptions only where needed.

2.  Use a **formatting tool** to automatically implement a consistent and readable code format.

    ```{r results="asis", echo=FALSE}
guide_table(
    r = htmltools::a("styler", href = "https://styler.r-lib.org/"),
    python = htmltools::a("Autopep8", href = "https://pypi.org/project/autopep8/"),
    julia = htmltools::a("JuliaFormatter.jl", href = "https://domluna.github.io/JuliaFormatter.jl/")
)
    ```

3.  Use a **style checking** tool to enforce a consistent and readable code style.

    ```{r results="asis", echo=FALSE}
guide_table(
    r = htmltools::a("lintr", href = "https://lintr.r-lib.org/"),
    python = htmltools::a("Pylint", href = "https://pypi.org/project/pylint/"),
    julia = htmltools::p("Enabled in VScode via", htmltools::a("StaticLint.jl", href = "https://github.com/julia-vscode/StaticLint.jl"))
)
    ```

## Life cycle

Life cycle management is simplified by reducing dependencies, and should comprise a central code repository.

1.  **Reduce dependencies** to simplify maintenance of your own package. Only depend on other packages that you trust and deem stable enough for the purpose, in order to avoid reinventing the wheel.

2.  Give clear information to users about changes in the package API via maintaining the **change log** and first **deprecating functionality** before removing it.

    ```{r results="asis", echo=FALSE}
guide_table(
    r = htmltools::p(
        htmltools::a("lifecycle", href = "https://cran.r-project.org/web/packages/lifecycle/index.html"),
        htmltools::a("fledge", href = "https://fledge.cynkra.com/")
    ),
    python = htmltools::a("deprecation", href = "https://deprecation.readthedocs.io/"),
    julia = htmltools::a("workflow", href = "https://invenia.github.io/blog/2022/06/17/deprecating-in-julia/")
)
    ```

3.  Use a **central repository** for version control, collecting and resolving issues, and managing releases. Include the publication of a **contributing guide** to help onboard new developers and enable community contributions.

# References

